Use of significance test logic by scientists in a novel reasoning task

### 2020-05-20 inst/manuscript/manuscript.Rmd
enabled font switching for base and ggplot fonts
moved to RSOS format

---
title: Use of significance test logic by scientists in a novel reasoning task
titlerunning: Statistical reasoning in scientists
authorrunning: Morey & Hoekstra
authors: 
- name: Richard D. Morey
  address: School of Psychology, Cardiff University
  email: moreyr@cardiff.ac.uk
- name: Rink Hoekstra
  address: Faculty of Behavioural and Social Sciences, University of Groningen
  email: r.hoekstra@rug.nl
keywords:
- cognitive science
- psychology
- science
- statistics
- statistical cognition
#PACS:

MSC:
- 91E10    
abstract: |
     Although statistical significance testing is one of the most widely-used techniques across science, previous research has suggested that scientists have a poor understanding of how it works. If scientists misunderstand one of their primary inferential tools the implications are dramatic: potentially unchecked, unjustified conclusions and wasted resources. Scientists' apparent difficulties with significance testing have led to calls for its abandonment or increased reliance on alternative tools, which would represent a substantial, untested, shift in scientific practice. However, if scientists' understanding of significance testing is truly as poor as thought, one could argue such drastic action is required. We show using a novel experimental method that scientists do, in fact, understand the logic of significance testing and can use it effectively. This suggests that scientists may not be as statistically-challenged as often believed, and that reforms should take this into account.
bibliography: "`r system.file('bib/bibfile.bib', package = 'MoreyHoekstra2019')`"
csl: nature.csl
output: rticles::springer_article
vignette: >
  %\VignetteIndexEntry{Manuscript}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
#nocite: | 
#  @Venables:Ripley:2002
header-includes: |
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{subfig}
  \renewcommand{\thesubfigure}{\Alph{subfigure}}
pkgdown:
  as_is: true
  extension: pdf
---

@@

knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE,
                      message = FALSE,
                      out.width = "0.95\\textwidth")

@@

Despite statistical significance testing (SST) use in a majority of research projects across fields, there is debate over whether scientists understand SST and can use it competently. Methodologists and statistical cognition researchers point to evidence from questionnaires and vignette studies to argue that researchers do not, in fact, grasp the core logic of SST. In one highly influential study of research psychologists, Oakes [-@Oakes:1986] presented six statements about a hypothetical significance test result to be categorized as true or false (e.g., "[The $p$ value provides] the probability of the null hypothesis being true"). Despite all of these statements being false, 97% of the research psychologists categorized at least one as true. Oakes argues that this shows that the participants have an "[un]sound understanding of the logic of the significance test" (p. 82).

@@

```{r interface, fig.cap="The interface on which samples were reported. The $x$-axis monotonically (but nonlinearly) related to the strength of the statistical evidence ($z$ statistic) favoring one group; the $y$-axis is monotonically (but nonlinearly) related to the sample size. Underlying numerical values of the statistical evidence and sample sizes were unknown to the participant."}
fn = system.file("pkg_html/img/clean_interface.png", package = "MoreyHoekstra2019")
knitr::include_graphics(fn)

Participants were also offered the opportunity to sample "random shuffle reports" that were described as the results of experiments with random assignment of elves to groups: that is, the result of experiments in which the null hypothesis was true. These results took no time to return. Participants were not told how to use these samples, only that they might use them.

@@

```{r error_rates, fig.cap="Decision error rates as a function of true effect size. The red solid line represents the rate at which participants {\\em correctly} determined the direction of the effect, except for at an effect size of 0, where it represents an incorrect claim of an effect. The shaded region around the solid line is one standard error wide. The difference between the black and the gray dashed lines shows the proportion of participants who decided that the two groups were the same (as opposed to failing to detect a difference)."}
MoreyHoekstra2019::fig_error_rates()
```

Decision error rates as a function of true effect size are shown in Figure \ref{fig:error_rates}.

@@

```{r response_ev, fig.cap="Statistical evidence underlying participants' decisions. The Wilcoxon rank-sum test $p$ value ($x$ axis) between the participants' null samples and experimental samples is used as a rough index of evidential strength. Kernel density estimates for the evidence are shown for three relevant conclusions. Each point at the bottom represents a single participant. Filled circles show correct decisions; hollow circles, incorrect decisions. The two hollow squares show sign errors."}
MoreyHoekstra2019::fig_response_evidence()
```

Figure \ref{fig:response_ev} shows the distribution of Wilcoxon $p$ values (arranged by the direction of the decision). Kernel density estimates show the distributions of $p$ values when participants indicated that Sparklies were faster, no detection/same, or that Jinglies were faster. With few exceptions, participants' conclusions appear reasonable given the information in the display, though a few participants appear to ignore clear evidence of an effect.

@@

```{r strategy_table}
MoreyHoekstra2019::christmas_stats_participants %>% 
  mutate(text_strong = text_comparison | text_sampling_var,
               text_only_weak = (text_asymmetry | text_inc_asymmetry ) & !text_strong,
               text_none = !(text_strong | text_only_weak),
               text_missing = text_missing | text_irrelevant) %>%
  summarise(Strong = sum(text_strong),
            `Only weak` = sum(text_only_weak),
            Neither = sum(text_none),
            `No shuffles` = sum(text_no_shuffle),
            `Missing` = sum(text_missing),
            Total = n()) -> for_table

@@

```{r response_ev2, fig.height=4, fig.cap="The effect of the evidence transformation manipulation on responding. Points on top (narrow scale; $q=7$) and bottom (wide scale; $q=3$) represent participants' decisions as a function of the most extreme experiment sampled. ``N'' indicates a ``no detect'' or ``same'' response; ``J/S'' indicates a response in favor of a difference between the groups. Curves show predicted probability by a logistic regression fit with standard errors.", fig.show = "hold", fig.subcap = c("Predicted response probabilities relative to visual extremity. Vertical lines show the critical 0.05 for the corresponding null sampling distribution.","Predicted response probabilities relative to the null sampling distributions (implicit $p$ values)."), out.width = "0.90\\textwidth", fig.ncol = 1}
glm_obj_x <- MoreyHoekstra2019::fig_evidence_scale_logistic_x()

@@

Figure \ref{fig:response_ev2} (top) shows responses (no detect/same or Jinglies/Sparklies) as a function of the most extreme experiment sampled ($x$ axis) and the transformation. There was a strong effect of the transformation consistent with use of the null sampling distribution; participants randomly assigned to the "narrow" evidence transformation responded "Jinglies/Sparklies" for much less visually extreme evidence (sequential LRT: $\chi^2_2 = `r round(sum(glm_anova_x[3:4,3]),3)`, p<.001$).

@@

Applying the same analysis to the responses corrected for their respective sampling distributions (Figure \ref{fig:response_ev2}, bottom) almost completely eliminates the effect of experimental condition, as would be expected if most participants were using the sampling distributions to calibrate (sequential LRT: $\chi^2_2 = `r round(sum(glm_anova_p[3:4,3]),3)`, p=`r round(1 - pchisq(sum(glm_anova_p[3:4,3]),2),3)`$). It is noteworthy that when they responses are aligned by sampling distribution, the wide condition appears to slightly dominate; this is consistent with a small number of participants incorrectly using the nondiagnostic visual extremeness to perform the task. If more people had been fooled by the irrelevant visual manipulation of the sampling distribution, we would expect this effect to be substantially larger.

@@

# References

<div id="refs"></div>

# End notes

* **Data and materials availability**: All materials, code, and data are freely available at https://github.com/richarddmorey/Morey_Hoekstra_StatCognition.

* **Author Contributions**: RDM conceptualized and designed the study in consultation with RH. RDM analysed the data and created the materials and figures. The manuscript was written by RDM and RH.

* The authors declare no competing interests.

* Correspondence and requests for materials should be addressed to Richard D. Morey (richarddmorey@gmail.com).

>>>>

---
title: Use of significance test logic by scientists in a novel reasoning task
author:
  - name: Richard D. Morey
    affiliation: "1"
  - name: Rink Hoekstra
    affiliation: "2"
address:
  - code: "1"
    address: School of Psychology, Cardiff University
  - code: "2"
    address: Faculty of Behavioural and Social Sciences, University of Groningen
corresp_author_name:  "Richard D. Morey"
corresp_author_email: "moreyr@cardiff.ac.uk"

subject:
  - "subject 1"
  - "subject 2"
  - "subject 3"

keywords:
  - reasoning
  - statistics
  - statistical cognition
  - decision making
  - significance testing

abstract: |
  Although statistical significance testing is one of the most widely-used techniques across science, previous research has suggested that scientists have a poor understanding of how it works. If scientists misunderstand one of their primary inferential tools the implications are dramatic: potentially unchecked, unjustified conclusions and wasted resources. Scientists' apparent difficulties with significance testing have led to calls for its abandonment or increased reliance on alternative tools, which would represent a substantial, untested, shift in scientific practice. However, if scientists' understanding of significance testing is truly as poor as thought, one could argue such drastic action is required. We show using a novel experimental method that scientists do, in fact, understand the logic of significance testing and can use it effectively. This suggests that scientists may not be as statistically-challenged as often believed, and that reforms should take this into account.

ethics: |
  This research project was evaluated by the Cardiff University School of Psychology (application number EC.18.12.11.5526G). It was found to be within the ethical guidelines for experiments with human participants. All participants gave informed consent prior to their participation.

data_accessibility: |
  All materials, code, and data are freely available at https://github.com/richarddmorey/Morey_Hoekstra_StatCognition.

author_contributions: |
  RDM conceptualized and designed the study in consultation with RH. RDM analysed the data and created the materials and figures. The manuscript was written by RDM and RH.

conflict_of_interest: |
  The authors declare no conflicts of interest.

bibliography: "`r system.file('bib/bibfile.bib', package = 'MoreyHoekstra2019')`"

lineno: true

site: bookdown::bookdown_site
output: 
  bookdown::pdf_book:
    base_format: rticles::rsos_article
pkgdown:
  as_is: true
  extension: pdf
vignette: >
  %\VignetteIndexEntry{Manuscript}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
header-includes: |
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{subfig}
  \renewcommand{\thesubfigure}{\Alph{subfigure}}
---

@@

knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE,
                      message = FALSE,
                      out.width = "0.95\\textwidth",
                      dev = "cairo_pdf",
                      fig.width = 6,
                      fig.height = 4)

@@

Despite statistical significance testing (SST) use in a majority of research projects across fields, there is debate over whether scientists understand SST and can use it competently. Methodologists and statistical cognition researchers point to evidence from questionnaires and vignette studies to argue that researchers do not, in fact, grasp the core logic of SST. In one highly influential study of research psychologists, Oakes [@Oakes:1986] presented six statements about a hypothetical significance test result to be categorized as true or false (e.g., "[The $p$ value provides] the probability of the null hypothesis being true"). Despite all of these statements being false, 97% of the research psychologists categorized at least one as true. Oakes argues that this shows that the participants have an "[un]sound understanding of the logic of the significance test" (p. 82).

\EndFirstPage

@@

```{r interface, fig.cap="The interface on which samples were reported. The $x$-axis monotonically (but nonlinearly) related to the strength of the statistical evidence ($z$ statistic) favoring one group; the $y$-axis is monotonically (but nonlinearly) related to the sample size. Underlying numerical values of the statistical evidence and sample sizes were unknown to the participant.", out.width=".9\\columnwidth"}
fn = system.file("pkg_html/img/clean_interface.png", package = "MoreyHoekstra2019")
knitr::include_graphics(fn)

```
Participants were also offered the opportunity to sample "random shuffle reports" that were described as the results of experiments with random assignment of elves to groups: that is, the result of experiments in which the null hypothesis was true. These results took no time to return. Participants were not told how to use these samples, only that they might use them.

@@

```{r error-rates, fig.cap="Decision error rates as a function of true effect size. The red solid line represents the rate at which participants {\\em correctly} determined the direction of the effect, except for at an effect size of 0, where it represents an incorrect claim of an effect. The shaded region around the solid line is one standard error wide. The difference between the black and the gray dashed lines shows the proportion of participants who decided that the two groups were the same (as opposed to failing to detect a difference).", fig.showtext=FALSE, out.width=".9\\columnwidth"}
MoreyHoekstra2019::fig_error_rates()
```

Decision error rates as a function of true effect size are shown in Figure \ref{fig:error-rates}.

@@

```{r response-ev, fig.cap="Statistical evidence underlying participants' decisions. The Wilcoxon rank-sum test $p$ value ($x$ axis) between the participants' null samples and experimental samples is used as a rough index of evidential strength. Kernel density estimates for the evidence are shown for three relevant conclusions. Each point at the bottom represents a single participant. Filled circles show correct decisions; hollow circles, incorrect decisions. The two hollow squares show sign errors.", out.width=".9\\columnwidth"}
MoreyHoekstra2019::fig_response_evidence()
```

Figure \ref{fig:response-ev} shows the distribution of Wilcoxon $p$ values (arranged by the direction of the decision). Kernel density estimates show the distributions of $p$ values when participants indicated that Sparklies were faster, no detection/same, or that Jinglies were faster. With few exceptions, participants' conclusions appear reasonable given the information in the display, though a few participants appear to ignore clear evidence of an effect.

@@

```{r strategy-table}
MoreyHoekstra2019::christmas_stats_participants %>% 
  mutate(text_strong = text_comparison | text_sampling_var,
               text_only_weak = (text_asymmetry | text_inc_asymmetry ) & !text_strong,
               text_none = !(text_strong | text_only_weak),
               text_missing = text_missing | text_irrelevant) %>%
  summarise(Strong = sum(text_strong),
            `Only weak` = sum(text_only_weak),
            Neither = sum(text_none),
            `No shuffles` = sum(text_no_shuffle),
            `Missing` = sum(text_missing),
            Total = n()) -> for_table

@@

```{r response-ev2, fig.height=4, fig.cap="The effect of the evidence transformation manipulation on responding. Points on top (narrow scale; $q=7$) and bottom (wide scale; $q=3$) represent participants' decisions as a function of the most extreme experiment sampled. ``N'' indicates a ``no detect'' or ``same'' response; ``J/S'' indicates a response in favor of a difference between the groups. Curves show predicted probability by a logistic regression fit with standard errors.", fig.show = "hold", fig.subcap = c("Predicted response probabilities relative to visual extremity. Vertical lines show the critical 0.05 for the corresponding null sampling distribution.","Predicted response probabilities relative to the null sampling distributions (implicit $p$ values)."), out.width=".9\\columnwidth", fig.ncol = 1}
glm_obj_x <- MoreyHoekstra2019::fig_evidence_scale_logistic_x()

@@

Figure \ref{fig:response-ev2} (top) shows responses (no detect/same or Jinglies/Sparklies) as a function of the most extreme experiment sampled ($x$ axis) and the transformation. There was a strong effect of the transformation consistent with use of the null sampling distribution; participants randomly assigned to the "narrow" evidence transformation responded "Jinglies/Sparklies" for much less visually extreme evidence (sequential LRT: $\chi^2_2 = `r round(sum(glm_anova_x[3:4,3]),3)`, p<.001$).

@@

Applying the same analysis to the responses corrected for their respective sampling distributions (Figure \ref{fig:response-ev2}, bottom) almost completely eliminates the effect of experimental condition, as would be expected if most participants were using the sampling distributions to calibrate (sequential LRT: $\chi^2_2 = `r round(sum(glm_anova_p[3:4,3]),3)`, p=`r round(1 - pchisq(sum(glm_anova_p[3:4,3]),2),3)`$). It is noteworthy that when they responses are aligned by sampling distribution, the wide condition appears to slightly dominate; this is consistent with a small number of participants incorrectly using the nondiagnostic visual extremeness to perform the task. If more people had been fooled by the irrelevant visual manipulation of the sampling distribution, we would expect this effect to be substantially larger.

@@
